How to Avoid OOTA Without Really Trying

### 2024-09-03 paper/ootavacuous.tex
asplos: Remove Section 7.2 ("Constraints of the Standard")
Instead, provide a brutal summary at the end of Section 4.6
("Volatile and Quasi Volatile Accesses").

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Section~\ref{sec:Constraints of the Standard}
presents examples illustrating some of these requirements.

@@

\subsection{Constraints of the Standard}
\label{sec:Constraints of the Standard}

\emph{Can we summarize this section with a list of permitted and prohibited
optimizations in the discussion of quasi-volatile?}

Volatile atomic accesses constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950} and
must be executed in strict accordance with the rules of the C++ abstract
machine.
This means that a volatile atomic store in the source code
corresponds directly to a machine-code store instruction, which can
commit only after all loads whose values are used to compute the
store's address and value have committed.
OOTA aficionados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

Although C++ nonvolatile accesses to atomic objects are not observable
behavior, any compiler that restricts its code analysis to a
single thread must assume (unless it can prove otherwise) that a given
relaxed atomic store can affect observable behavior in a different
thread that loads the value stored.
This assumption does not constrain the compiler to anywhere near
the extent that a volatile relaxed atomic store would, but it does add
significant constraints over those related to nonvolatile non-atomic
stores.

For example, nonvolatile relaxed atomic accesses are subject to the
C++ memory model.
The ``order-preserving'' bullet point from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering of quasi-volatile accesses when they are
to the same object,
but such reordering is in any case directly forbidden by the memory model
(the four coherence rules discussed in 6.9.2.2p14--19 \co{[intro.races]}).
The ``onto'' bullet point also largely reiterates restrictions that follow
from the standard, in particular,
that compilers should not invent atomic stores,
duplicate atomic stores, or invent atomic loads.
On the other hand, it is permissible for compilers to omit redundant
atomic stores or fuse nonvolatile atomic accesses of adjacent objects
under appropriate circumstances;
these topics correspond to the ``total'' and ``one-to-one'' bullet points of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

\paragraph{Invented Atomic Stores}
\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/invented-store.litmus @@
\caption{Example Invented Store}
\label{lst:Example Invented Store}
\end{listing}
The reason that atomic stores should not be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors that are
forbidden by the abstract machine.
To see this, consider Listing~\ref{lst:Example Invented Store},
a \co{herd7} litmus test that demonstrates such a behavior.
Without P0's line~1, only the values~0 and~3 can be loaded into \co{r1} by
P1's line~1.
With P0's line~1, the additional value~42 can also be loaded into \co{r1}.
The compiler is therefore forbidden from inventing the store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors.
For example, the compiler might be able to prove that there
are no other accesses to \co{x} at the time of the invented
store---but it's very hard to imagine how a compiler that analyzes
only a single thread at a time could prove such a thing.

\paragraph{Duplicated Atomic Stores}
\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/duplicated-store.litmus @@
\caption{Example Duplicated Store}
\label{lst:Example Duplicated Store}
\end{listing}
Duplicating atomic stores is a special case of inventing them, but it is
worth illustrating the fact that duplicating stores can also introduce
new and undesirable behaviors.
To see this, consider Listing~\ref{lst:Example Duplicated Store}, a
litmus test that demonstrates such a behavior.
Without P0's line~1, if the value loaded into \co{r1} is one then the final value
of \co{x} must be two.
With that line, the final value of \co{x} can be one even when the
value loaded into \co{r1} is one.
The compiler is therefore forbidden from duplicating atomic stores
unless it can prove that doing so does not negatively affect the
program's observable behaviors.
Again, the compiler might be able to prove that there
are no other accesses to \co{x} at the time of the duplicated store.

In contrast, compilers should not duplicate, repurpose, or otherwise
invent atomic loads, as discussed in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back nonvolatile atomic stores to the
same object
always might be executed such that no other thread accesses the object
during the time between those two stores.
This means that if the compiler omitted the first store,
the user would have no way to prove this fact short of
inspecting the assembly code.
In cases where such omissions are undesirable, the user can resort to
volatile atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent them.

Omitting a redundant store cannot create an OOTA cycle,
because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is a valid execution of the
program with the store present.
% This is awkward because a reader might reasonably extend the "Acting
% as if" all the way to the comma.  The best I could do was "It is not
% possible to create an OOTA cycle by acting as if a redundant store
% was omitted from the source code", which is admittedly not all that
% great.

\paragraph{Fused Accesses of Adjacent Nonvolatile Atomic Objects}
Suppose a pair of adjacent nonvolatile atomic objects, when combined,
form a single machine-word aligned and machine-word sized unit.
Suppose further that the source code contains a pair of atomic
loads, one from each of these objects, with the last of the pair
being a relaxed load.
Then the compiler could generate instead
a single atomic load of the combined pair,
with the memory-ordering semantics of the first load.
A similar fact applies to pairs of stores.

This holds even when the atomic objects are treated as quasi volatile,
providing an example of how the ``one-to-one'' bullet point of
Section~\ref{sec:Volatile and Quasi Volatile Accesses} does not apply
to them.

>>>>

Applying these requirements to specific examples results in prohibiting
invention or duplication of atomic loads and stores.
However, in the absence of ordering constraints, it permits omitting
redundant non-volatile atomic stores and fusing of non-volatile atomic
accesses to adjacent objects.

@@
